Density-dependent selection and the limits of relative fitness

### 2018-08-30 manuscript.tex
l

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations. These conditions are met in most previous models of density-dependent selection (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions. We develop a novel model of density-regulated population growth with three ecologically intuitive traits: fecundity, mortality, and competitive ability. Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

@@

\subsection*{Density-regulating traits and the threat of strong selection}

For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection. Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection. Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing. This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait threatens Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models. This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}. We consider the simple birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density). This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}). The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to threaten the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

@@

The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e. as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the density-independent approximation, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig. 9.5 in \cite{otto_2011}; Fig. 5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

@@

Suppose that we naively used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$. Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, $\sum_j \langle x_j\rangle_a \approx l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$. Thus, the rare type would be predicted to decline in frequency even though it has identical traits. This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type. To avoid this, the effective means must be taken in a way that ensures that the total propagule density $\langle X \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

For the focal type $i$, we use the conditional dispersal distributions $p$ to calculate the effective mean, 
\begin{equation}
\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}. \label{eq:effectivefocal}
\end{equation}
However, for nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$. Specifically, we assume that $X$ follows a  Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$.

@@

In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium to equilibrium sweep of a type with higher $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

>>>>

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations. These conditions are met in most previous models of density-dependent selection (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions. We develop a novel model of density-regulated population growth with three ecologically intuitive traits: fecundity, mortality, and competitive ability. Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium pose the most serious difficulties for relative fitness models. In such cases our model offers a possible alternative to relative fitness.

@@

\subsection*{Density-regulating traits under strong selection}

For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection. Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection. Thus, to invalidate Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing. This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can break down even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for the $b$ trait in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait can invalidate Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models. This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}. We consider the simple birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

@@

Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density). This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}). The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}. The combination of both is necessary to invalidate the constant-$s$ approximation. Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.

@@

The above findings suggest that the largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment. While transient deviations from demographic equilibrium driven by the appearance of new types can also cause the density-independent approximation to break down, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold). By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g. Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.

@@

Suppose that we naively used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$. Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, $\sum_j \langle x_j\rangle_a \approx l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$. Thus, the rare type would be predicted to decline in frequency even though it has identical traits.

This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical. The effective means must thus be taken in a way that ensures that the expected total propagule density is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.

We achieve this as follows. For nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$. Specifically, we assume that $X$ follows a Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$. However, for the focal type $i$, we use the exact conditional dispersal distributions $p$ to calculate the effective mean, 
\begin{equation}
\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}. \label{eq:effectivefocal}
\end{equation}
This assumption is mathematically convenient

@@

In section ``Density-regulating traits under strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium to equilibrium sweep of a type with higher $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.
